---
title: "Bias Test"
author: "Water BSU"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_caption: true
    theme: united
    highlight: tango
    css: |
      .centered {
        text-align: center;
      }
---

```{r setup, echo = FALSE,message=FALSE, results='hide'}
suppressWarnings({
library(knitr)
library(tidyverse)
library(sf)
library(ggplot2)
library(SimDesign)
library(transport)
library(gridExtra)
library(heatmaply)
library(maps)
library(sp)
library(leaflet)
library(rmarkdown)
})
```
```{r, echo = FALSE}
test_variable <- "pre_mm_cyr"
date <- Sys.Date()
```


<div class="centered">
***Variables: `r test_variable`     |     `r date`***
</div>




```{r, echo=FALSE,results='hide',fig.keep='all',fig.cap="Figure 1: [Wasserstein Distance](https://www.sciencedirect.com/topics/mathematics/wasserstein-distance) of the 13 selected variables."}
suppressWarnings({
# importing data
data <- read.csv('H:/Final_data.csv', header=TRUE)

# importing names associated with geospatial variables and major freshwater habitat type
VARnames <- read.csv('C:/Users/stevenschmitz/Desktop/PlacementBias/Git/G4-master/in/VARnames.csv', header=FALSE)
ECOnames <- read.csv('C:/Users/stevenschmitz/Desktop/PlacementBias/Git/G4-master/in/ECOnames.csv', header=FALSE)

# --------------------------------------------------------------------------------------------
# Calculating standard bias and Wassenstein distance for gauge reaches vs. all reaches

# sub-setting reaches to those containing gauges and removing reaches (n=22) with missing geospatial data
gagdata<-data %>% 
  filter(!is.na(Gage_No)) %>%
  select(COMID,uparea,order_,dor_pc_pva,slope,tmp_dc_cyr,pre_mm_cyr,crp_pc_use,urb_pc_use,pac_pc_cse,ppd_pk_uav,hft_ix_u09,gdp_ud_usu,ecoregion) 
# selecting all reaches and removing reaches (n=13,143) with missing missing geospatial data
alldata<-data %>% 
  filter(!is.na(dor_pc_pva)) %>%
  select(COMID,uparea,order_,dor_pc_pva,slope,tmp_dc_cyr,pre_mm_cyr,crp_pc_use,urb_pc_use,pac_pc_cse,ppd_pk_uav,hft_ix_u09,gdp_ud_usu,ecoregion) 

# calculating variable means for all data

varmeans<-alldata[,-1] %>%
  summarise_all(mean,na.rm=TRUE)

# calculating standardized bias and Wasserstein distance for each variable
all_bias<-matrix(, nrow = dim(gagdata[,-1])[2], ncol = 3)
rownames(all_bias)<-t(VARnames)
all_bias<-cbind(VARnames,all_bias)
all_bias[,2]<-bias(gagdata[,-1],varmeans,type='standardized')
colnames(all_bias)<-c("Variable", "bias", "wasser","Direction")

# note that the first column of gagdata and alldata is omitted b/c it is COMID
for (p in 1:dim(gagdata[, -1])[2]) {
  gagdata_std <- (gagdata[, p + 1] - mean(alldata[, p + 1])) / sd(alldata[, p + 1])
  all_bias[p, 3] <- wasserstein1d(gagdata_std, scale(alldata[, p + 1]), p = 1)
  
  # Check for missing values before making the comparison
  if (!is.na(all_bias[p, 2]) && all_bias[p, 2] > 0) {
    all_bias[p, 4] <- "positive"
  } else {
    all_bias[p, 4] <- "negative"
  }
}

# --------------------------------------------------------------------------------------------
# Producing Figure 2

# Figure 2a
all_bias <- all_bias[order(all_bias$wasser), ] 
all_bias$Variable <- factor(all_bias$Variable, levels = all_bias$Variable)
all_bias$Direction <- factor(all_bias$Direction, levels = c("positive","negative"))

ggplot(all_bias, aes(x=Variable, y=`wasser`, color=Direction, size=wasser, fill=Direction)) + 
  geom_point(alpha=1)  + scale_size(range = c(1, 6)) +
  scale_color_manual(values=c("blue", "red"),name="Bias Direction", labels = c("Positive","Negative")) + 
  scale_fill_manual(values=c("blue", "red"),name="Bias Direction", labels = c("Positive","Negative")) + 
  labs(title="a",y="Wasserstein Distance (Bias)") + 
  coord_flip() +
  guides(size="none") +
  scale_y_continuous(limits=c(0,0.8,0)) +
  theme (
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(colour="black",size=12), 
    axis.text.y = element_text(colour="black",size=12),
    legend.text = element_text(colour="black",size=11),
    legend.title=element_text(colour="black",size=13), 
    plot.title=element_text(colour="black",size=14,hjust=0), 
    axis.title.y=element_blank(), 
    axis.title.x=element_text(colour="black",size=12), 
    panel.background = element_rect(fill = "lightgray", colour="black"),
    legend.position = c(0.825, 0.1)
  )
})
```




```{r, echo=FALSE,results='hide',fig.keep='all', fig.cap="Figure 2: variable distributions showing cumulative probabilities of variables occurrence."}
suppressWarnings({
temp<-c(rep("gag",dim(gagdata)[1]),rep("all",dim(alldata)[1]))
temp<-as.matrix(temp)
comdata<-cbind(temp,as.data.frame(rbind(gagdata,alldata)))
colnames(comdata)[1]<-"type"

a<-ggplot(comdata, aes(x = pre_mm_cyr,colour=type)) + stat_ecdf(size=1.5) +
  scale_color_manual(values=c("black","#E69F00"),name="River Segment", labels = c("All","Gauged")) + 
  labs(title="a", x="Precipitation (mm/yr)", y="Cumulative probability")  +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = c(0.85, 0.25),
    panel.background = element_rect(fill = "lightgray"),
    axis.text.x = element_text(colour="black",size=11), 
    axis.text.y = element_text(colour="black",size=11),
    legend.text = element_text(colour="black",size=10),
    legend.title=element_text(colour="black",size=10), 
    #plot.title=element_text(colour="black",size=14,hjust=0), 
    axis.title.y=element_text(colour="black",size=11), 
    axis.title.x=element_text(colour="black",size=11), 
  )

b<-ggplot(comdata, aes(x = log10(dor_pc_pva+1),colour=type)) + stat_ecdf(size=1.5) +
  theme_bw() + 
  scale_color_manual(values=c("black","#E69F00")) +
  theme(panel.grid.minor = element_blank(),legend.position = "none",panel.background = element_rect(fill = "lightgray")) +
  labs(title="b", x="Flow Regulation (log+1) (%)", y="Cumulative probability")  +
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "lightgray"),
    axis.text.x = element_text(colour="black",size=11), 
    axis.text.y = element_text(colour="black",size=11),
    axis.title.y=element_text(colour="black",size=11), 
    axis.title.x=element_text(colour="black",size=11), 
  )

c<-ggplot(comdata, aes(x = hft_ix_u09,colour=type)) + stat_ecdf(size=1.5) +
  theme_bw() + 
  scale_color_manual(values=c("black","#E69F00")) +
  theme(panel.grid.minor = element_blank(),legend.position = "none",panel.background = element_rect(fill = "lightgray")) +
  labs(title="c", x="Human Footprint", y="Cumulative probability")  +
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "lightgray"),
    axis.text.x = element_text(colour="black",size=11), 
    axis.text.y = element_text(colour="black",size=11),
    axis.title.y=element_blank(), 
    axis.title.x=element_text(colour="black",size=11), 
  )

d<-ggplot(comdata, aes(x = tmp_dc_cyr/10,colour=type)) + stat_ecdf(size=1.5) +
  theme_bw() + 
  scale_color_manual(values=c("black","#E69F00")) +
  theme(panel.grid.minor = element_blank(),legend.position = "none",panel.background = element_rect(fill = "lightgray")) +
  labs(title="d", x="Air Temperature (°C)", y="Cumulative probability")   +
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "lightgray"),
    axis.text.x = element_text(colour="black",size=11), 
    axis.text.y = element_text(colour="black",size=11),
    axis.title.y=element_blank(), 
    axis.title.x=element_text(colour="black",size=11), 
  )

# plotting in a 2-by-2 panel
grid.arrange(a,c,b,d)
})

# Producing Figure S4

# transforming data to aid intepretation
comdata$type <- factor(comdata$type, levels = c("all","gag"))
comdata$uparea_log<-log(comdata$uparea+1)
comdata$slope_log<-log10(comdata$slope+1)
comdata$pre_mm_cyr_log<-log10(comdata$pre_mm_cyr+1)
comdata$crp_pc_use_log<-log10(comdata$crp_pc_use+1)
comdata$urb_pc_use_log<-log10(comdata$urb_pc_use+1)
comdata$pac_pc_cse_log<-log10(comdata$pac_pc_cse+1)
comdata$ppd_pk_uav_log<-log10(comdata$ppd_pk_uav+0.01)
comdata$gdp_ud_usu_log<-log10(comdata$gdp_ud_usu+1)

a<-ggplot(comdata, aes(x = uparea_log,colour=type)) + stat_ecdf(size=1.5)  +
  theme_bw() + 
  scale_color_manual(values=c("black", "#E69F00"),name="River Segment", labels = c("All", "Gauged"))  +
  labs(title="a", x=expression(Catchment~Area~(log+1)~(m^2)), y="Cumulative probability")  +
  scale_y_continuous(limits=c(0,1.0),labels = scales::number_format(accuracy = 0.01)) +
  theme(
    legend.key.size = unit(0.3, 'cm'),
    legend.title = element_text(size=10),
    panel.grid.minor = element_blank(),
    legend.position = c(0.7, 0.28),
    panel.background = element_rect(fill = "lightgray")
  )
b<-ggplot(comdata, aes(x = order_,colour=type)) + stat_ecdf(size=1.5)  +
  theme_bw() + 
  scale_y_continuous(limits=c(0,1.0),labels = scales::number_format(accuracy = 0.01)) +
  scale_color_manual(values=c("black","#E69F00")) +
  theme(panel.grid.minor = element_blank(),legend.position = "none",panel.background = element_rect(fill = "lightgray"),axis.title.y=element_blank()) +
  labs(title="b", x="Stream Order (Strahler)")  
c<-ggplot(comdata, aes(x = slope_log,colour=type)) + stat_ecdf(size=1.5)  +
  theme_bw() + 
  scale_color_manual(values=c("black","#E69F00")) +
  scale_y_continuous(limits=c(0,1.0),labels = scales::number_format(accuracy = 0.01)) +
  theme(panel.grid.minor = element_blank(),legend.position = "none",panel.background = element_rect(fill = "lightgray"),axis.title.y=element_blank()) +
  labs(title="c", x=expression(Channel~Gradient~(log)~"(%)"))  
d<-ggplot(comdata, aes(x = pre_mm_cyr_log,colour=type)) + stat_ecdf(size=1.5)  +
  theme_bw() + 
  scale_color_manual(values=c("black","#E69F00")) +
  scale_y_continuous(limits=c(0,1.0),labels = scales::number_format(accuracy = 0.01)) +
  theme(panel.grid.minor = element_blank(),legend.position = "none",panel.background = element_rect(fill = "lightgray")) +
  labs(title="d", x=expression(Precipitation~(log+1)~"(mm/year)"), y="Cumulative probability")
e<-ggplot(comdata, aes(x = crp_pc_use_log,colour=type)) + stat_ecdf(size=1.5)  +
  theme_bw() + 
  scale_color_manual(values=c("black","#E69F00")) +
  scale_y_continuous(limits=c(0,1.0),labels = scales::number_format(accuracy = 0.01)) +
  theme(panel.grid.minor = element_blank(),legend.position = "none",panel.background = element_rect(fill = "lightgray"),axis.title.y=element_blank()) +
  labs(title="e", x="Crop Landuse (log+1) (%)")
f<-ggplot(comdata, aes(x = urb_pc_use_log,colour=type)) + stat_ecdf(size=1.5)  +
  theme_bw() + 
  scale_color_manual(values=c("black","#E69F00")) +
  scale_y_continuous(limits=c(0,1.0),labels = scales::number_format(accuracy = 0.01)) +
  theme(panel.grid.minor = element_blank(),legend.position = "none",panel.background = element_rect(fill = "lightgray"),axis.title.y=element_blank()) +
  labs(title="f", x="Urban Landuse (log+1) (%)")
g<-ggplot(comdata, aes(x = pac_pc_cse_log,colour=type)) + stat_ecdf(size=1.5)  +
  theme_bw() + 
  scale_color_manual(values=c("black","#E69F00")) +
  scale_y_continuous(limits=c(0,1.0),labels = scales::number_format(accuracy = 0.01)) +
  theme(panel.grid.minor = element_blank(),legend.position = "none",panel.background = element_rect(fill = "lightgray")) +
  labs(title="g", x="Protected Area (log+1) (%)", y="Cumulative probability")
h<-ggplot(comdata, aes(x = ppd_pk_uav_log,colour=type)) + stat_ecdf(size=1.5)  +
  theme_bw() + 
  scale_color_manual(values=c("black","#E69F00")) +
  scale_y_continuous(limits=c(0,1.0),labels = scales::number_format(accuracy = 0.01)) +
  theme(panel.grid.minor = element_blank(),legend.position = "none",panel.background = element_rect(fill = "lightgray"),axis.title.y=element_blank()) +
  labs(title="h", x=expression(Pop~Density~(log+0.01)~(no.~km^-2)))  
i<-ggplot(comdata, aes(x = gdp_ud_usu_log,colour=type)) + stat_ecdf(size=1.5)  +
  theme_bw() + 
  scale_color_manual(values=c("black","#E69F00")) +
  scale_y_continuous(limits=c(0,1.0),labels = scales::number_format(accuracy = 0.01)) +
  theme(panel.grid.minor = element_blank(),legend.position = "none",panel.background = element_rect(fill = "lightgray"),axis.title.y=element_blank()) +
  labs(title="i", x="Gross Domest Prod (log+1) (USD)")  

# plotting in a 4-by-4 panel
grid.arrange(a,b,c,d,e,f,g,h,i)
```

<div class="centered">
***Installing New Gage***
</div>

The next step is selecting the GRADES river segment that will create the largest reduction in bias for the select variable(s) by creating a 'simulated gage.' Bias calculations are reworked with a 'phantom' gage assigned to each successive ungaged segment in each loop. The segment that produces the largest reduction in bias to the select variable(s) is highlighted below in red with adjacent river sections. Each GRADES segment has an associated COMID that identifies segment length, the parent river section, and geospatial information.


```{r, echo=FALSE,results='hide',fig.keep='all', fig.cap="Figure 3: GRADES river segment that minimizes variable(s) bias plotted with adjacent segments and geospatial information."}
data <- read.csv('H:/Final_data.csv', header=TRUE)

# importing names associated with geospatial variables and major freshwater habitat type
VARnames <- read.csv('C:/Users/stevenschmitz/Desktop/PlacementBias/Git/G4-master/in/VARnames_noeco.csv', header=FALSE)
FHTnames <- read.csv('C:/Users/stevenschmitz/Desktop/PlacementBias/Git/G4-master/in/FHTnames.csv', header=FALSE)
ECOnames <- read.csv('C:/Users/stevenschmitz/Desktop/PlacementBias/Git/G4-master/in/ECOnames.csv', header=FALSE)

# sub-setting reaches to those containing gauges and removing reaches (n=22) with missing geospatial data
gagdata<-data %>% 
  filter(!is.na(Gage_No)) %>%
  select(COMID,uparea,order_,dor_pc_pva,slope,tmp_dc_cyr,pre_mm_cyr,crp_pc_use,urb_pc_use,pac_pc_cse,ppd_pk_uav,hft_ix_u09,gdp_ud_usu) 
# selecting all reaches and removing reaches (n=13,143) with missing missing geospatial data
alldata<-data %>% 
  filter(!is.na(dor_pc_pva)) %>%
  select(COMID,uparea,order_,dor_pc_pva,slope,tmp_dc_cyr,pre_mm_cyr,crp_pc_use,urb_pc_use,pac_pc_cse,ppd_pk_uav,hft_ix_u09,gdp_ud_usu) 

# calculating variable means for all data

varmeans<-alldata[,-1] %>%
  summarise_all(mean,na.rm=TRUE)

# defining the number of segments
no.seg<-dim(alldata)[1]
permutation_bias<-matrix(,ncol=13,nrow=no.seg)
for (j in 1:no.seg) {
  perdata<-rbind(gagdata, alldata[j,])
  permutation_bias[j,]<-c(alldata[j,1],bias(perdata[,2:(dim(gagdata)[2])],varmeans,type='standardized'))
}

# calculating % change in bias for each variable and overall mean across variables
current_bias<-as.matrix(bias(gagdata[,2:(dim(gagdata)[2])],varmeans,type='standardized'))
temp<-sweep(permutation_bias[,2:13], MARGIN=2,FUN="-", current_bias)
finalbias<-sweep(temp,MARGIN=2,FUN="/", current_bias)*100
finalbias<-cbind(permutation_bias[,1],finalbias,rowMeans(finalbias[,2:12]))
finalbias<-as.data.frame(finalbias)
colnames(finalbias)<-c("COMID", t(VARnames), "MeanBiasChange")

#This function takes the 'variable_name' argument as a string, refer to alldata import variables
# variable name - variable you are testing bias for
# alldata - see above
#
# This function is telling you, based on all of the locations in alldata, where the bias reduction
# would be the most if a gage were placed there.
# 
# output: COMID:river segment where, if gage were placed, results in max bias reduction
# 

place_bias <- function(variable_name, alldata, gagdata, varmeans) {
  if (!(variable_name %in% colnames(gagdata))) stop(paste("Variable", variable_name, "not found in gagdata"))
  
  variables <- c("uparea", "order_", "dor_pc_pva", "slope", "tmp_dc_cyr",
                 "pre_mm_cyr", "crp_pc_use", "urb_pc_use", "pac_pc_cse",
                 "ppd_pk_uav", "hft_ix_u09", "gdp_ud_usu")
  
  if (!(variable_name %in% variables)) stop(paste("Variable", variable_name, "not found in the common variables"))
  
  result <- data.frame(COMID = numeric(0), Position = numeric(0), BiasReduction = numeric(0))
  
  for (j in 1:nrow(alldata)) {
    if (j %in% gagdata$Position) next
    
    bias_before <- bias(gagdata[, variables], varmeans, type = 'standardized')
    perdata <- rbind(gagdata, alldata[j, ])
    bias_after <- bias(perdata[, variables], varmeans, type = 'standardized')
    
    variable_index <- which(names(bias_before) == variable_name)
    reduction <- bias_before[variable_index] - bias_after[variable_index]
    
    result <- rbind(result, data.frame(COMID = alldata[j, "COMID"], Position = j, BiasReduction = reduction))
  }
  
  max_reduction_row <- result[which.max(result$BiasReduction), ]
  return(max_reduction_row)
}

biasoutput <- place_bias("tmp_dc_cyr", alldata, gagdata, varmeans) # output saved to variable

#Plot grades segment with lowest bias

shapefile_path <- 'H:/GRADES/GRADES_eco.shp'
grades_sf <- st_read(shapefile_path)
comid <- biasoutput$COMID
selected_row <- grades_sf[grades_sf$COMID == comid, ]

ggplot() +
  geom_sf(data = grades_sf) +
  geom_sf(data = selected_row, color = "red") +
  labs(title = paste("COMID =", comid), subtitle = paste("Segment Location: ",round(st_bbox(selected_row)$ymax,3),",",round(st_bbox(selected_row)$xmax,3))) +
  coord_sf(xlim = c(st_bbox(selected_row)$xmin - 0.1, st_bbox(selected_row)$xmax + 0.1),
           ylim = c(st_bbox(selected_row)$ymin - 0.1, st_bbox(selected_row)$ymax + 0.1)) 
```



```{r leaflet, echo=FALSE,fig.keep='all',results='asis',fig.cap="Figure 4: GRADES river segment overlaid on satellite imagery"}
# leaflet
compute_median_coordinates <- function(multilinestring) {
  coords <- st_coordinates(multilinestring)
  median_coords <- apply(coords, 2, median)
  return(median_coords)
}

median_coordinates <- compute_median_coordinates(selected_row$geometry)
median_coordinates[1] <- round(median_coordinates[1],4)
median_coordinates[2] <- round(median_coordinates[2],4)
median_coordinates <- median_coordinates[1:2]
median_coords_str <- paste(median_coordinates, collapse = ", ")

leaflet() %>%
  addPolylines(data = selected_row, color = "red", weight = 4, group = "Selected Line") %>%
  addMarkers(lat = median_coordinates[2], lng = median_coordinates[1],
             label = paste("Segment location", median_coords_str), labelOptions = labelOptions(noHide = TRUE)) %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  addProviderTiles(providers$Esri.WorldImageryLabels, group = "Labels") %>%
  addLayersControl(
    baseGroups = c("Satellite"),
    overlayGroups = c("Labels", "Selected Line"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

